---
title: GREGOR functional enrichment analysis of SuSiE and mvSuSiE fine-mapping results for UK Biobank blood cell traits
author: Peter Carbonetto
date: November 23, 2023
output:
  html_document:
    theme: readable
    lib_dir: site_libs
    self_contained: no
---

Here we generate some plots for the paper from the GREGOR functional enrichment analysis performed [here](https://github.com/gaow/mvarbvs/blob/master/analysis/20231106_GREGOR_functional_enrichment.ipynb).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in this analysis.

```{r load-pkgs, message=FALSE, warning=FALSE}
library(ggplot2)
library(cowplot)
```

These are the blood cell traits.

```{r traits}
traits <- c("RBC","HGB","MCV","RDW","MSCV",          # mature red blood cell
            "Reticulocyte_perc","HLR_perc",          # immature red blood cell
            "Platelet_count","Plateletcrit","PDW",   # platelet
            "WBC","Lymphocyte_perc","Monocyte_perc", # white blood cell
            "Neutrophill_perc","Eosinophill_perc","Basophill_perc",
            "crosstrait")
```

These are the SNP annotations we will show the enrichment results for.

```{r annotations}
annotations <- list(
  "transcription factor binding sites" =
    c("TFBS_ENCODE.bed","FOSL2.bed","FOXO1_DeMayo.bed",
	  "NR2F2.bed","PLZF.bed","POLII_DeMayo.bed"),
  "enhancer promoter regions" =
    c("Enhancer_Andersson.bed","Enhancer_Hoffman.bed",
      "PromoterFlanking_Hoffman.bed","Promoter_UCSC.bed",
	  "SuperEnhancer_Hnisz.bed","TSS_Hoffman.bed",
	  "WeakEnhancer_Hoffman.bed"),
  "eqtls data" =
    c("eQTL_0.05_FDR.bed","eQTL_0.1_FDR.bed"),
  "genomic structural elements" =
    c("Coding_UCSC.bed","Intron_UCSC.bed","UTR_3_UCSC.bed","UTR_5_UCSC.bed"))
```

Import the SuSiE-based and mvSuSiE-based enrichment results from the
files and combine into a single data frame.

```{r import-results}
odds_ratios <- NULL
for (trait in traits) {
  susie_file <- paste("../output/blood_cell_traits/gregor/SuSiE",trait,
                      "positive_cs.index_gregor_output_StatisticSummaryFile",
                      "fisher.txt",sep = "_")
  mvsusie_file <- paste("../output/blood_cell_traits/gregor/mvSuSiE",trait,
                        "positive_cs.index_gregor_output_StatisticSummaryFile",
                        "fisher.txt",sep = "_")
  susie   <- read.table(susie_file,sep = " ",header = TRUE,
                        stringsAsFactors = FALSE)
  mvsusie <- read.table(mvsusie_file,sep = " ",header = TRUE,
                        stringsAsFactors = FALSE)
  odds_ratios <- 
    rbind(odds_ratios,
	      data.frame(bed       = susie$Bed_File,
	                 trait     = trait,
	                 method    = "susie",
					 oddsratio = susie$odds,
					 stringsAsFactors = FALSE),
		  data.frame(bed       = mvsusie$Bed_File,
		             trait     = trait,
  		             method    = "mvsusie",
		             oddsratio = mvsusie$odds,
		  stringsAsFactors = FALSE))
}
```

Prepare the data frame for ggplot2.

```{r prepare-results}
n <- length(annotations)
for (i in 1:n) {
  x <- subset(odds_ratios,
              method == "mvsusie" &
			  trait == "crosstrait" &
              is.element(bed,annotations[[i]]))
  rows <- order(x$oddsratio,decreasing = TRUE)
  annotations[[i]] <- x[rows,"bed"]
}
annotations <- unlist(annotations)
odds_ratios <- subset(odds_ratios,is.element(bed,annotations))
odds_ratios <- transform(odds_ratios,
                         bed    = factor(bed,rev(annotations)),
                         trait  = factor(trait,traits),
						 method = factor(method,c("susie","mvsusie")))
```

This first plot shows the "cross-trait" SuSiE and mvSuSiE enrichment
results:

```{r plot-crosstrait, fig.height=3.5, fig.width=4.5}
p1 <- ggplot(subset(odds_ratios,trait == "crosstrait"),
       aes(x = oddsratio,y = bed,color = method)) +
  geom_point() +
  scale_color_manual(values = c("dodgerblue","red")) +
  labs(y = "") +
  theme_cowplot(font_size = 9) +
  theme(panel.grid.major.y = element_line(color = "gray",size = 0.3))
p1
```

Next we show the enrichment results for all traits:

```{r plot-all-traits, fig.height=9, fig.width=8}
p2 <- ggplot(odds_ratios,aes(x = oddsratio,y = bed,color = method)) +
  geom_point() +
  scale_color_manual(values = c("dodgerblue","red")) +
  labs(y = "") +
  theme_cowplot(font_size = 7) +
  facet_wrap(vars(trait)) +
  theme(strip.background = element_blank(),
        panel.grid.major.y = element_line(color = "gray",size = 0.3))
p2
```


```{r ggsave, include=FALSE}
ggsave("../plots/gregor_global.eps",p1,height = 3.5,width = 4.25)
```
